Generalists are the most urban-tolerant of birds: an analysis of ecological and life history traits using a novel continuous urbanization index

### 2018-06-25 ms.Rmd
changes for Proc B.

---
title: "Generalists are the most urban-tolerant of birds: a phylogenetically controlled analysis of ecological and life-history traits using a novel continuous urbanization index"

author: |
  | Corey T. Callaghan^[Corresponding author: c.callaghan@unsw.edu.au] $^1$, Richard E. Major $^1$$^,$$^2$, John H. Wilshire $^1$, John M. Martin $^1$$^,$$^3$, Richard T. Kingsford $^1$, William K. Cornwell $^1$$^,$$^4$
  | $^1$Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, NSW, Australia; $^2$Australian Museum Research Institute, Sydney, NSW, Australia; $^3$Royal Botanic Gardens and Domain Trust, Sydney, NSW, Australia; $^4$Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, NSW, Australia
date: ""
output:
  rmarkdown::pdf_document:
    fig_caption: yes    
    includes:  
      in_header: preamble-latex.tex
fontsize: 12pt
link-citations: no
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ecology-letters.csl
bibliography: references.bib
---

@@

@@

We aimed to determine what suite of life history traits predicted avian tolerance to urbanization at a continental scale. We predicted that different traits would positively and negatively associate with urbanization. For generality, our analyses had to (1) be comprehensive in faunal coverage, (2) account for phylogenetic constraints, (3) include a broad-range of traits, and (4) employ a continuous measure of urbanization. We integrated a series of large datasets to establish a novel continuous urbanization index to test associations with ecological and life history traits influencing species' responses to urbanization. We first investigated whether our urbanization index was statistically independent in relation to phylogeny (i.e., phylogenetic signal - Revell et al. 2008; Evans et al. 2011), given our comprehensive sample size. We then developed models, including and excluding phylogenetic constraints, to test for associations between our urbanization index and a suite of continuous and categorical predictor variables representing species' traits.

@@

We filtered bird observations [@callaghan2017assessing; @la2014role], only including observations on complete checklists and which were either travelling, random, stationary, area, or followed BirdLife Australia protocols [@barrett2003new] (see here: http://help.ebird.org/customer/portal/articles/1006209-understanding-observation-types). We then filtered the checklists to those which travelled < 5 km or covered < 500 Ha, to ensure the highest quality data were used. Any checklists shared among multiple observers were subsampled, including only one checklist to avoid duplication. This provided 637,482 eligible checklists throughout Australia. Seabirds (e.g., gannets, petrels, shearwaters) were omitted from analyses as we did not expect any correlation with seabirds using urban areas. Further, only species with > 100 observations were analyzed. In total, 580 species, most (~80%) of Australia's bird species were candidates for analyses. We used differing taxonomic authorities (see Appendix 2).

@@

## Ecological and life history traits
Ecological and life history traits were chosen on the basis of existing support (e.g., McClure 1989; Kark *et al.* 2007; Evans *et al.* 2011; Table 1; Appendix 1). These included a wide array of potential traits for all of Australia's birds [@garnett2015biological]. A total of 22 traits were included in analyses: clutch size, body size, brain residual, nest generalism, feeding habitat generalism, breeding habitat generalism, diet generalism, habitat association with agricultural landscapes, habitat association with treed/forested habitats, habitat association with grass/shrubland habitats, whether a species nests in hollows, whether a species nests on the ground, whether a species eats plants, whether a species is a granivore, whether a species is an insectivore, whether a species eats carrion, whether a species is nomadic/irruptive, whether a species is migratory, the range size of a species, nesting aggregation of a species, feeding aggregation of a species, and whether a species is a cooperative breeder (Table 1; Appendix 1a). For any missing trait in the dataset [@garnett2015biological] for a species, we assumed it did not exist. For example, if a species had no data for a specific habitat, it was deemed not to occur in that habitat. Our preliminary analyses, which included IUCN status, showed that exotic birds strongly associated with urbanization; these were omitted from the final analyses to avoid bias, given the preferential selection by acclimatization societies for their proven synanthropy [@blackburn2009avian]. We also excluded IUCN status because most (96%) of our candidate species were of Least Concern. In total, 477 species had complete data for all candidate traits and were included in analyses.

@@

## Phylogenetic analyses
### Phylogenetic signal
There was a distinct relationship between the phylogenetic signal and our urbanization index (Fig. 3; Fig. S3; Table 2), significantly different from random for 4 out of 5 indices (Cmean, I, K.star, and Lambda). For example, the urbanization index was disproportionately important with respect to the phylogenetic tree (Fig. 3), evidenced by the highest urbanization indices clustered throughout the waterbird clade. Further, within a specific clade, there were some species which substantially higher urbanization indices (e.g., *Manorina melanocephala*, Noisy Miner, cf. honeyeaters). We generally found weaker signals for behavioral, compared with morphological traits. For behavioural traits, the estimates of phylogenetic signal were weaker than Brownian motion, but usually significant (Table 2). Additionally, there was a strong phylogenetic signal for body size and brain residual size (Table 2).

@@

## Acknowledgments
We thank the > 6,000 eBirders who have contributed data in Australia. We also thank the dedicated eBird team at the Cornell Lab of Ornithology and the global network of reviewers. Thanks to Stephen Garnett and Glenn Ehmke for providing us with extent of occurrence estimates for each of the species in our analysis. Thanks to Corrine Edwards for providing the illustrations used in Fig. 3.

@@

# Tables
\small
```{r Table 2, echo=FALSE, warning=FALSE, message=FALSE}
library(readr)
Table1a <- read_csv("Data/Table1.csv")
colnames(Table1a) <- gsub("X1", "Variable", colnames(Table1a))
is.num <- sapply(Table1a, is.numeric)
Table1a[is.num] <- lapply(Table1a[is.num], round, 3)

Table1b <- read_csv("Data/TableS2.csv")
colnames(Table1b) <- gsub("X1", "Variable", colnames(Table1b))
is.num <- sapply(Table1b, is.numeric)
Table1b[is.num] <- lapply(Table1b[is.num], round, 3)

grouped <- rbind(cbind(Value="Statistic", Table1a), cbind(Value="p-values", Table1b))
tgrp <- xtabs(cbind(Cmean, I, K, K.star, Lambda) ~ Value+Variable, grouped)

@@

library(kableExtra)
library(pander)
pander(table, caption = "Summary of results showing an assessment of a phylogenetic signal for our response variable (urbanization index) and the continuous predictor traits. We show all available indices (Blomberg's K and K*, Abouheif's Cmean, Moran's I, and Pagel's Lambda; Keck et al. 2016), each of which represents the statistical nonindependence of a trait's value due to its phylogenetic relatedness (Revell et al. 2008).", split.table=Inf)

@@

collinearity_investigation <- read_csv("Data/TableS1.csv")
collinearity_investigation <- collinearity_investigation[-1,]
collinearity_investigation <- collinearity_investigation %>%
  rename(Trait = X1)

>>>>

---
title: "Generalists are the most urban-tolerant of birds: an analysis of ecological and life history traits using a novel continuous urbanization index"

author: |
  | Corey T. Callaghan^[Corresponding author: c.callaghan@unsw.edu.au] $^1$, Richard E. Major $^1$$^,$$^2$, John H. Wilshire $^1$, John M. Martin $^1$$^,$$^3$, Richard T. Kingsford $^1$, William K. Cornwell $^1$$^,$$^4$
  | $^1$Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, NSW, Australia; $^2$Australian Museum Research Institute, Sydney, NSW, Australia; $^3$Royal Botanic Gardens and Domain Trust, Sydney, NSW, Australia; $^4$Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, NSW, Australia
date: ""
output:
  rmarkdown::pdf_document:
    fig_caption: yes    
    includes:  
      in_header: preamble-latex.tex
fontsize: 12pt
link-citations: no
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/proceedings-of-the-royal-society-b.csl
bibliography: references.bib
---

@@

*Keywords*: eBird, citizen science, phylogenetics, life history traits, avian, brain size

@@

We aimed to determine what suite of life history traits predicted avian tolerance to urbanization at a continental scale. We predicted that different traits would positively and negatively associate with urbanization. For generality, our analyses had to (1) be comprehensive in faunal coverage, (2) account for phylogenetic constraints, (3) include a broad-range of traits, and (4) employ a continuous measure of urbanization. We integrated a series of large datasets to establish a novel continuous urbanization index to test associations with ecological and life history traits influencing species' responses to urbanization. We first investigated whether our urbanization index was statistically independent in relation to phylogeny (i.e., phylogenetic signal - [@revell2008phylogenetic; @evans2011makes]), given our comprehensive sample size. We then developed models, including and excluding phylogenetic effects, to test for associations between our urbanization index and a suite of continuous and categorical predictor variables representing species' traits.

@@

We filtered bird observations [@callaghan2017assessing; @la2014role], only including observations on complete checklists and which were either travelling, random, stationary, area, or followed BirdLife Australia protocols [@barrett2003new] (see here: http://help.ebird.org/customer/portal/articles/1006209-understanding-observation-types). We then filtered the checklists to those which travelled < 5 km or covered < 500 Ha, to ensure the highest quality data were used. Any checklists shared among multiple observers were subsampled, including only one checklist to avoid duplication. This provided 637,482 eligible checklists throughout Australia. Seabirds (e.g., gannets, petrels, shearwaters) were omitted from analyses as we did not expect any correlation with seabirds using urban areas. Further, only species with > 100 observations were analyzed. In total 580 species, most (~80%) of Australia's bird species, were candidates for analyses. We used differing taxonomic authorities (see Appendix 2).

@@

## Ecological and life history traits
Ecological and life history traits were chosen on the basis of existing support [e.g., @mcclure1989characterizes; @kark2007living; @evans2011makes] (Table 1; Appendix 1), including a wide array of potential traits for all of Australia's birds [@garnett2015biological]. A total of 22 traits, grouped into 7 categories, were included (Table 1): life history traits (n=3), measures of niche breadth (n=4), specific habitat preferences (n=3), specific nesting preferences (n=2), specific diet preferences (n=4), movement and range size (n=3), and degree of sociality (n=3). For any missing trait in the dataset [@garnett2015biological] for a species, we assumed it did not exist. For example, if a species had no data for a specific habitat, it was deemed not to occur in that habitat. Our preliminary analyses, which included IUCN status, showed that exotic birds strongly associated with urbanization; these were omitted from the final analyses to avoid bias, given the preferential selection by acclimatization societies for their proven synanthropy [@blackburn2009avian]. We also excluded IUCN status because most (96%) of our candidate species were of Least Concern. In total, 477 species had complete data for all candidate traits and were included in analyses.

@@

## Phylogenetic analyses
### Phylogenetic signal
There was a distinct relationship between the phylogenetic signal and our urbanization index (Fig. 3; Fig. S3; Table 2), significantly different from random for 4 out of 5 indices (Cmean, I, K.star, and Lambda). For example, the urbanization index was disproportionately important with respect to the phylogenetic tree (Fig. 3), evidenced by the highest urbanization indices clustered throughout the waterbird clade. Further, within a specific clade, there were some species with substantially higher urbanization indices (e.g., *Manorina melanocephala*, Noisy Miner, cf. honeyeaters). We generally found weaker signals for behavioral, compared with morphological traits. For behavioural traits, the estimates of phylogenetic signal were weaker than Brownian motion, but usually significant (Table 2). Additionally, there was a strong phylogenetic signal for body size and brain residual size (Table 2).

@@

# Data, code and materials
This work relied heavily on open-source, fully accessible datasets.
The necessary portions of those datasets, along with code to reproduce the analyses are available at
this github repository (https://github.com/cornwell-lab-unsw/bird_urbanness), and will be
permanently archived upon acceptance of the manuscript. The supporting data has been included in the article's supplementary material (Appendix 4).

# Competing interests
We have no competing interests.

# Author's Contributions
CTC, WKC, and JHW conceived and designed the methodology of the
study. CTC, WKC performed the statistical analyses. JMM, REM, and RTK contributed in drafting
the manuscript. All authors provided editorial advice, approved the final version of the manuscript,
and agree to be accountable for all aspects of the work.

# Acknowledgments
We thank the > 6,000 eBirders who have contributed data in Australia. We also thank the dedicated eBird team at the Cornell Lab of Ornithology and the global network of reviewers. Thanks to Stephen Garnett and Glenn Ehmke for providing us with extent of occurrence estimates for each of the species in our analysis. Thanks to Corrine Edwards for providing the illustrations used in Fig. 3.

# Funding
The authors declare no sources of funding specific to this work.

@@

# Tables
\small
```{r Table 2, echo=FALSE, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
Table1a <- read_csv("Data/Table1.csv")
colnames(Table1a) <- gsub("X1", "Variable", colnames(Table1a))
is.num <- sapply(Table1a, is.numeric)
Table1a[is.num] <- lapply(Table1a[is.num], round, 3)
Table1a <- Table1a %>%
  rename(`Abouheif's Cmean` = Cmean) %>%
  rename(`Moran's I` = I) %>%
  rename(`Blomberg's K` = K) %>%
  rename(`K*` = K.star) %>%
  rename(`Pagel's Lambda` = Lambda)

Table1b <- read_csv("Data/TableS2.csv")
colnames(Table1b) <- gsub("X1", "Variable", colnames(Table1b))
is.num <- sapply(Table1b, is.numeric)
Table1b[is.num] <- lapply(Table1b[is.num], round, 3)
Table1b <- Table1b %>%
  rename(`Abouheif's Cmean` = Cmean) %>%
  rename(`Moran's I` = I) %>%
  rename(`Blomberg's K` = K) %>%
  rename(`K*` = K.star) %>%
  rename(`Pagel's Lambda` = Lambda)

grouped <- rbind(cbind(Value="Statistic", Table1a), cbind(Value="p-values", Table1b))
tgrp <- xtabs(cbind(`Abouheif's Cmean`, `Moran's I`, `Blomberg's K`, `K*`, `Pagel's Lambda`) ~ Value+Variable, grouped)

@@

library(kableExtra)
library(pander)
pander(table, caption = "Summary of results showing an assessment of a phylogenetic signal for our response variable (urbanization index) and the continuous predictor traits. We show all available indices (Keck et al. 2016), each of which represents the statistical nonindependence of a trait's value due to its phylogenetic relatedness (Revell et al. 2008).", split.table=Inf)

@@

collinearity_investigation <- read_csv("Data/TableS1.csv")
collinearity_investigation <- collinearity_investigation[-1,]
collinearity_investigation <- collinearity_investigation %>%
  rename(Trait = X1)
collinearity_investigation <- collinearity_investigation[c(2, 1, 4, 16, 3, 6, 11, 5, 22, 20, 15, 14, 10, 7, 8, 9, 13, 12, 21, 18, 19, 17), ]
